HO unification from object language to meta language

### 2024-03-20 paper/main.tex
rework 2.1

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$, $\sigma \subseteq \sigma'$ when $\sigma$ is more general than $\sigma'$,
and we assume that the unification of our meta language is correct:
%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

@@

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

@@

\noindent
Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
terms) and a call to \solve{} on the set of links. We claim the following:

@@

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
\end{array}
$$

@@

\subsection{The intuition in a nutshell}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where any
``problematic'' subterm $tp$ is replaced by a unification variable $h$
and an accessory link, that represent a suspended unification problem
$h \Ue tp$. As a result \Ue is well behaved on $t$, that is it captures \Eo.
We now define ``problematic'' formally.

\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
\begin{definition}[\maybeeta]
  $\maybeeta = \{ t | \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

@@

\begin{definition}[\maybebeta]
  $\maybebeta = \{ X t_1 \ldots t_n | t \not\in \llambda \}$.
\end{definition}

@@

\marginpar{TODO: find better word}
\newcommand{\normal}{\ensuremath{\mathrm{normal}}}
\begin{definition}[Normal form]
Given a set of terms $X \subseteq \Ho{}$,
$$
\normal(X) =
\mathcal{P}(X) \cap (\maybebeta{} ~\cup~ \maybeeta{}) = \emptyset
$$
\end{definition}

\noindent
We write $\sigma X = \{ \sigma t | t \in X\}$.

\begin{proposition}[Normal form preservation]\label{prop:nf}
$\forall \mathcal{T}, \forall \mathbb{L}, \forall p, \forall \sigma, \forall \sigma'$
$$
\normal(\sigma\mathcal{T}) \land
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
\Rightarrow \normal((\sigma \cup \sigma')\mathcal{T})
$$
\end{proposition}

\noindent
In particular this guarantees that is we start from normal terms
we never introduce eta-long or non-beta-normal terms in $\sigma'$.

Note that proposition \ref{prop:nf} does not hold for \Uo{} since
decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta.

>>>>

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
with substitution $\sigma$; we write $\sigma t$ for the application of
the substitution to $t$, and $\sigma X = \{ \sigma t | t \in X\}$ when
$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$. We assume that the unification of our meta
language is correct:
%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

@@

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{check}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

@@

\noindent
Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
terms) and a call to \progress{} on the set of links. We claim the following:

@@

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
\end{array}
$$

@@

\subsection{The intuition in a nutshell}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where any
``problematic'' subterm $tp$ is replaced by a fresh unification variable $h$
and an accessory link that represent a suspended unification problem
$h \Ue tp$. As a result \Ue is ``well behaved'' on $t$, meaning it does not
contradict \Eo (as it would do on ``problematic'' terms).
We now define ``problematic'' and ``well behaved'' more formally.

\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
\begin{definition}[\maybeeta]
  $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

@@

\begin{definition}[\maybebeta]
  $\maybebeta = \{ X t_1 \ldots t_n ~|~ X t_1 \ldots t_n \not\in \llambda \}$.
\end{definition}

@@

\newcommand{\wellb}{\ensuremath{\mathcal{W}}}
\begin{definition}[Well behaved set]
Given a set of terms $X \subseteq \Ho{}$,
$$
\wellb(X) \Leftrightarrow \forall t \in \mathcal{P}(X), t \not\in (\maybebeta{} ~\cup~ \maybeeta{})
$$
\end{definition}

\noindent

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \mathcal{T}, \forall \mathbb{L}, \forall p, \forall \sigma, \forall \sigma'$
$$
\wellb(\sigma\mathcal{T}) \land
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
\Rightarrow \wellb((\sigma \cup \sigma')\mathcal{T})
$$
\end{proposition}

\noindent
A less formal way to state \ref{prop:nf} is that \hstep{} never
``commits'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
that could be erased by an $\eta$-contraction),
nor puts in $\sigma$ a flexible application outside \llambda{} (an application
node that could be erased by a $\beta$-reduction).

Note that proposition \ref{prop:nf} does not hold for \Uo{} since
decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta that were move out of the way (put in $\mathbb{L}$)
by compilation.
